---
title: "Introduction to sparsebn"
author: "Bryon Aragam"
date: "`r Sys.Date()`"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Introduction to sparsebn}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

`sparsebn` is an `R` package for learning sparse Bayesian networks and other graphical models from high-dimensional data via sparse regularization. Designed to handle:

- Experimental data with interventions 
- Mixed observational / experimental data
- High-dimensional data with _p >> n_
- Datasets with thousands of variables (tested up to _p_=8000)
- Continuous and discrete data

The workhorse behind [`sparsebn`](http://www.github.com/itsrainingdata/sparsebn/) is the [`sparsebnUtils`](http://www.github.com/itsrainingdata/sparsebnUtils/)
package, which provides various S3 classes and methods for representing and manipulating graphs. The basic algorithms are implemented in [`ccdrAlgorithm`](http://www.github.com/itsrainingdata/ccdrAlgorithm/) and [`discretecdAlgorithm`](http://www.github.com/gujyjean/discretecdAlgorithm/).

## Overview

The main methods for learning graphical models are:

* `estimate.dag` for directed acyclic graphs (Bayesian networks).
* `estimate.precision` for undirected graphs (Markov random fields).
* `estimate.covariance` for covariance matrices.

Currently, estimation of precision and covariances matrices is limited to Gaussian data.

## Introduction: Structure learning in five lines of code

The following example illustrates how to use `sparsebn` to learn a Bayesian network in just a few lines of code. An explanation of this code can be found in the next section.

```{r message=FALSE, warning=FALSE}
library(sparsebn)
data(pathfinder)
data <- sparsebnData(pathfinder[["data"]], type = "continuous")
dags <- estimate.dag(data)
dags
```

## Example: Learning the pathfinder network

In this section, we will reconstruct the pathfinder network from the [Bayesian network repository](http://www.bnlearn.com/bnrepository/#pathfinder). The pathfinder network has 109 nodes and 195 edges.

In order to use the methods in the sparsebn package, we need to indicate what kind of data we are working with by wrapping the data into a `sparsebnData` object. First we load the data and then create a `sparsebnData` object:

```{r}
data(pathfinder)
dat <- sparsebnData(pathfinder$data, type = "continuous", ivn = NULL)
```

The `sparsebnData` object keeps track of what kind of data we are working with: Is it discrete or continuous, does it contain an interventions, and if it is discrete, what are the factor levels in the data? The argument `type = "continuous"` specifies that this data is continuous, and `ivn = NULL` indicate that there are no interventions. 

Now we can run the algorithm:

```{r message=FALSE, warning=FALSE}
dags <- estimate.dag(data = dat)
dags
```

Instead of automatically generating a grid of regularization parameters, we can also generate one manually (see `?generate.lambdas`). 

```{r message=FALSE, warning=FALSE}
nn <- num.samples(dat) # number of samples in the dataset / equivalent to nrow(dat$data)
lambdas <- generate.lambdas(sqrt(nn), 0.05, lambdas.length = 50, scale = "linear")
dags <- estimate.dag(data = dat, 
                     lambdas = lambdas,
                     verbose = FALSE)
dags
```

Note that the output is a _solution path_ (stored internally as a `sparsebnPath` object), instead of a single estimate. In order to select a particular DAG, we need to do model selection. For example, we can visualize the solution with 195 edges:
```{r}
solution <- select(dags, edges = 195)
plotDAG(solution)
```

For comparison, let's plot the original pathfinder graph. 
```{r}
plotDAG(pathfinder$dag)
```

Alternatively, we can automatically select a good solution using `select.parameter`:
```{r}
select.idx <- select.parameter(dags, dat)
solution <- select(dags, index = select.idx)
plotDAG(solution)
```

The output of `estimate.dag` is a list of _graphs_, i.e. without weights. In order to do estimate the weights in these graphs, use `estimate.parameters`:
```{r}
out.fit <- estimate.parameters(dags, data = dat)
```

The output is a list of weights, one for each value of $\lambda$. The weights are given in terms of $(B,\Omega)$, corresponding to the list `list(coefs, vars)`. For example, we can see how the weight of the first node on the second changes as we decrease $\lambda$:
```{r}
unlist(lapply(out.fit, function(x) x$coefs[1,2]))
```

## Example: Learning a Markov Chain

In this example, we illustrate how to learn a simple Markov chain on three nodes. We will also discuss how to use `sparsebn` to do covariance estimation. Suppose that the data is generated by the following graphical model:

$$X_1\to X_2\to X_3.$$

Assume unit influences between variables, i.e. $X_j\sim\mathcal{N}(0,1)$ and $X_{j} = X_{j-1} + \varepsilon_j$ with $\varepsilon_j\sim\mathcal{N}(0,1)$ for $j>1$. If $X=(X_1,X_2,X_3)$ then $X=B^TX+\varepsilon\sim\mathcal{N}(0,\Sigma)$, where we use the following parameters:

$$
B = \begin{pmatrix}
0 & 1 & 0 \\
0 & 0 & 1 \\
0 & 0 & 0
\end{pmatrix}, \quad
\Omega = \begin{pmatrix}
1 & 0 & 0 \\
0 & 1 & 0 \\
0 & 0 & 1
\end{pmatrix}, \quad
\Sigma = \begin{pmatrix}
3 & 2 & 1 \\
2 & 2 & 1 \\
1 & 1 & 1
\end{pmatrix}.
$$

To generate data from this model, first define the covariance matrix:
```{r}
mean.vector <- rep(0, 3)
covariance.matrix <- rbind(c(3,2,1),
                           c(2,2,1),
                           c(1,1,1))
```

Then we can generate some data using `mvtnorm::rmvnorm`:
```{r}
gaussian.data <- mvtnorm::rmvnorm(n = 100, mean = mean.vector, sigma = covariance.matrix)
colnames(gaussian.data) <- c("X1", "X2", "X3")
```

Now we can use this data to estimate $B$:
```{r}
dat <- sparsebnData(gaussian.data, type = "continuous")
dags <- estimate.dag(data = dat, 
                         lambdas.length = 20, 
                         edge.threshold = 10, 
                         verbose = FALSE)
dags
```

As expected, the third estimate in our solution path gives the correct estimate:
```{r}
dags[[3]]
get.adjacency.matrix(dags[[3]])
```

We can also use this data to directly estimate the covariance matrix $\Sigma$:
```{r, warning=FALSE, message=FALSE}
cov.out <- estimate.covariance(data = dat)
```

Compared with the output of `estimate.dag`, which is a more complicated `sparsebnPath` object, the output of `estimate.covariance` (and also `estimate.precision`) is simply a list of matrices:
```{r}
class(cov.out)
```

Let's take a look at the third estimate in the solution path (corresponding to the correct estimate of $B$ from before):
```{r}
cov.out[[3]]
```

If we increase our sample size to $n=1000$, the estimate gets closer to the truth:
```{r, warning=FALSE, message=FALSE}
gaussian.data <- mvtnorm::rmvnorm(n = 1000, mean = mean.vector, sigma = covariance.matrix)
dat <- sparsebnData(gaussian.data, type = "continuous")
cov.out <- estimate.covariance(data = dat)
cov.out[[3]]
```

## Appendix: Simulating data from a linear Gaussian SEM

Both datasets above have been simulated from a linear Gaussian SEM. In this section, we illustrate how to do this from scratch.

We first need a DAG; for this example we will use the pathfinder network from the previous section. First, load this DAG:

```{r}
data(pathfinder)
B <- as.matrix(get.adjacency.matrix(pathfinder$dag)) # pathfinder network as an adjacency matrix
```

If $X=B^TX+\varepsilon\sim\mathcal{N}(0,\Sigma)$ and $\varepsilon\sim\mathcal{N}(0,\Omega)$, then one can show that:
$$
\Sigma = (I-B)^{-T}\Omega(I-B)^{-1}.
$$

Assuming unit influences (i.e. $\beta_{ij}=1$ if $\beta_{ij}\ne 0$) and unit variances (i.e. $\omega_j^2=1$ for all $j$), we can then compute $\Sigma$ directly:

```{r}
id <- diag(rep(1, num.nodes(pathfinder$dag)))   # 109x109 identity matrix
Omega <- id                                     # conditional variances
Sigma <- solve(t(id - B)) %*% Omega %*% solve(id - B)
```

Finally, we can use the `mvtnorm` package to generate random multivariate Gaussian data:

```{r}
set.seed(123)
nn <- 1000
gaussian.data <- mvtnorm::rmvnorm(nn, sigma = Sigma)
```

Instead of setting $\beta_{ij}=1$, we can also use random edge weights:
```{r}
B[B!=0] <- runif(n = num.edges(pathfinder$dag), min = 0.5, max = 2)
Sigma <- solve(t(id - B)) %*% Omega %*% solve(id - B)
gaussian.data <- mvtnorm::rmvnorm(nn, sigma = Sigma)
```
